Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-30 ind_files/06_nontechnical_discussion.tex
Rephrase after proofreading

Lastly, noise-induced selection is particular to fluctuating populations and does not occur in models with fixed population sizes\footnote{If $\sum_j x_j$ is a constant, the map $x_i \to x_i/\sum_j x_j$ becomes a linear map and we no longer need It\^o's formula to move from densities to frequencies in the derivation I conduct in Appendix \ref{App_density_to_freq}; Thus, simply dividing equation \eqref{nD_Ito_SDE} by the (now constant) total population size provides the complete dynamics of the system in frequency space: Note that the directional terms in equation \eqref{nD_Ito_SDE} depend only on $\mathbf{A}^-$, which in turn depends only on the fitness $w_i$ and the mutation terms, and this system thus has no noise-induced selection.} such as the neutral Wright-Fisher or Moran models, suggesting that working with such constant population frameworks is not sufficient to accurately capture the dynamics of real populations. Several theoretical studies have pointed out that the empirically common practice of approximating the size of fluctuating populations through a constant `effective population size' (often obtained as the harmonic mean of population size over time) is of limited applicability~\citep{gillespie_natural_1974,sjodin_meaning_2005,parsons_consequences_2010,iizuka_effective_2010,abu_awad_effects_2018,kuosmanen_turnover_2022}. Recent experimental evolution studies have also directly shown that the harmonic mean of population size need not be a good proxy for capturing and/or predicting evolutionary dynamics in fluctuating populations~\citep{chavhan_larger_2019}. In our case, noise-induced selection cannot be captured by any constant effective population size, harmonic mean or otherwise, since noise-induced selection effects also depend on the turnover $\tau_i$ of each type and are thus a property of each \emph{type} in a fluctuating population (and not merely a function of the population size as a whole), meaning that each type would demand its own effective population size for the entire population. My results thus lend further support to one of the general messages of these previous studies - Approximating fluctuating populations via a simpler constant (effective) population size requires some careful justification, since it may inadvertently remove important evolutionary properties of the systems under study.

>>>>

Lastly, noise-induced selection is particular to fluctuating populations and does not occur in models with fixed population sizes\footnote{If $\sum_j x_j$ is a constant, the map $x_i \to x_i/\sum_j x_j$ becomes a linear map and we no longer need It\^o's formula to move from densities to frequencies in the derivation I conduct in Appendix \ref{App_density_to_freq}; Thus, simply dividing equation \eqref{nD_Ito_SDE} by the (now constant) total population size provides the complete dynamics of the system in frequency space: Note that the directional terms in equation \eqref{nD_Ito_SDE} depend only on $\mathbf{A}^-$, which in turn depends only on the fitness $w_i$ and the mutation terms, and this system thus has no noise-induced selection.} such as Wright-Fisher or Moran models, suggesting that working with such constant population frameworks is not sufficient to accurately capture the dynamics of real populations. Several theoretical studies have pointed out that the empirically common practice of approximating the size of fluctuating populations through a constant `effective population size' (often obtained as the harmonic mean of population size over time) is of limited applicability~\citep{gillespie_natural_1974,sjodin_meaning_2005,parsons_consequences_2010,iizuka_effective_2010,abu_awad_effects_2018,kuosmanen_turnover_2022}. Recent experimental evolution studies have also directly shown that the harmonic mean of population size need not be a good proxy for capturing and/or predicting evolutionary dynamics in fluctuating populations~\citep{chavhan_larger_2019}. In our case, noise-induced selection cannot be captured by any constant effective population size, harmonic mean or otherwise, since noise-induced selection effects also depend on the turnover $\tau_i$ of each type and are thus a property of each \emph{type} in a fluctuating population (and not merely a function of the population size as a whole), meaning that each type would demand its own effective population size for the entire population. My results thus lend further support to one of the general messages of these previous studies - Approximating fluctuating populations via a simpler constant (effective) population size requires some careful justification, since it may inadvertently remove important evolutionary properties of the systems under study.

### 2023-03-30 ind_files/99_Appendix_D_examples.tex
Rephrase after proofreading

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Thus, we have the equations
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dW_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we obtain the logistic equation. This derivation also makes it clear that two systems with very different stochastic dynamics can nevertheless converge to the same infinite population limit \eqref{ex_1D_stoch_logistic_det_limit}, since equation \eqref{ex_1D_stoch_logistic_det_limit} only depends on the difference $\lambda-\mu$.
\myfig{0.6}{figures/App_stoch_logistic_same_w_different_t.png}{\textbf{Comparison of time series produced by Gillespie simulations of the stochastic logistic equation for two different parameter values.} Simulations with the parameter values $\lambda = 4, \mu = 2$ are plotted in red, and simulations with the parameter values $\lambda = 31, \mu = 29$ are plotted in blue. Both simulations have $K=200$. The infinite population limit of both simulations is plotted in black dotted lines. The graph shows 10 realizations each for the two sets of parameter values.}{fig_1D_same_w_different_t_comparison}
To illustrate the effects of this seemingly innocent fact, figure \ref{fig_1D_same_w_different_t_comparison} compares two simulations which have the same value of $\lambda - \mu$ but a ten-fold difference in $\lambda+\mu$. As is clear from the figure, though both populations have the same behavior at the infinite population limit, populations with a higher value of $\lambda+\mu$ exhibit much wilder fluctuations and are therefore more prone to stochastic extinction. This is the root cause of the noise-induced selection that occurs in higher dimensions, discussed in detail in part \ref{part_summary}.

>>>>

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $\mu + (\lambda-\mu)\frac{n}{K}$, where $\mu$ and $K$ are positive constants. Thus, we have the birth and death rates
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dW_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we obtain the logistic equation. This derivation also makes it clear that two systems with very different stochastic dynamics can nevertheless converge to the same infinite population limit \eqref{ex_1D_stoch_logistic_det_limit}, since equation \eqref{ex_1D_stoch_logistic_det_limit} only depends on the difference $\lambda-\mu$.
\myfig{0.6}{figures/App_stoch_logistic_same_w_different_t.png}{\textbf{Comparison of time series produced by Gillespie simulations of the stochastic logistic equation for two different parameter values.} Simulations with the parameter values $\lambda = 4, \mu = 2$ are plotted in red, and simulations with the parameter values $\lambda = 31, \mu = 29$ are plotted in blue. Both simulations have $K=200$. The infinite population limit of both simulations is plotted in black dotted lines. The graph shows 10 realizations each for the two sets of parameter values.}{fig_1D_same_w_different_t_comparison}
To illustrate the effects of this seemingly innocent fact, figure \ref{fig_1D_same_w_different_t_comparison} compares two simulations which have the same value of $\lambda - \mu$ but a ten-fold difference in $\lambda+\mu$. As is clear from the figure, though both populations have the same behavior at the infinite population limit, populations with a higher value of $\lambda+\mu$ exhibit much wilder fluctuations and are therefore more prone to stochastic extinction. This is the root cause of the noise-induced selection that occurs in higher dimensions, discussed in detail in part \ref{part_summary}.
